---
title: "ARIMA Modeling and Forecasting of the French Dairy Production Index"
author: "Varnel TIENTCHEU & Rivalien MAGUETSWET"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
---

This project focuses on the econometric modeling and forecasting of a French economic time series. Using the Box-Jenkins methodology, we aim to identify, estimate, and validate an optimal model to predict future values of industrial output.

## Part I: The Data

### a) Data description

**Sector and Definition**

The series selected for this study is the **Industrial Production Index (IPI)** for the **Manufacture of Dairy Products** (NAF rev. 2, group 10.5) in France. This index, identified by Insee ID **010767631**, tracks the monthly volume of output for the dairy transformation industry.

**Time Span and Observations**

-   **Period:** The dataset covers the period from **January 1990 to February 2026**.

-   **Frequency:** Monthly.

-   **Sample Size:** With 434 observations.

The series is obtained from the Insee time series databank and has undergone specific pre-processing:

-   **CVS-CJO (SA-WDA):** The data is corrected for **seasonal variations** and **working day effects**. This is critical as it removes deterministic seasonal patterns (like the spring "milk flush") and calendar-related noise, allowing us to focus on the underlying stochastic process.

-   **Logarithmic Transformation:** We apply a natural log transformation ($ln(X_t)$) to the raw data. This is justified by the need to:

    1.  **Stabilize Variance:** as the magnitude of fluctuations in industrial production often grows proportionally with the level of the index.

    2.  **Linearize the Model:** Prepare the data for linear $ARIMA$ modeling where additive relationships are assumed.

    ```{r echo=FALSE}
    library(readr)
    # Importation de la base
    base <- read_delim("valeurs_mensuelles.csv", delim = ";", skip = 3)

    colnames(base) <- c("t", "Xt", "status")

    # Conversion en Date (Format YYYY-MM)
    base$t <- as.Date(paste0(base$t, "-01"), format="%Y-%m-%d")
    base$Xt <- as.numeric(base$Xt)
    base$status <- as.character(base$status)

    # On réordonne la série
    base <- base[order(base$t), ]
    str(base)

    ```

### b) Stationnarity

```{r, echo=FALSE}
#install.packages("forecast")
library(forecast)

# Passage au logarithme
xt <- ts(base$Xt, start = c(1990, 1), frequency = 12)
lxt <- log(xt)

# Extraction du mouvement de fond 

trend <- ma(lxt, order = 12, centre = TRUE)

par(mfrow=c(1,2))
# Visualisation propre
plot(lxt, 
     main = "Evolution du Logarithme de l'IPI (Produits Laitiers)",
     ylab = "Log-Indice", 
     xlab = "Temps",
     col = "gray",
     lwd = 2)

lines(trend, col = "firebrick", lwd = 1) 

legend("bottomright", 
       legend = c("Log(Xt)", "Tendance"),
       col = c("gray", "firebrick"), 
       lty = 1, 
       lwd = c(1, 1))

# Visualisation de la série tedantanciée
plot(lxt-trend, 
     main = "Série détendanciée",
     ylab = "Xt-Trend", 
     xlab = "Temps",
     col = "blue",
     lwd = 1)
#mvt de fond de Xt-trend
tr <- ma(lxt-trend, order = 12, centre = TRUE)
lines(tr, col = "firebrick", lwd = 1)

legend("bottomright", 
       legend = c("Log(Xt)-trend", "Tendance"),
       col = c("gray", "firebrick"), 
       lty = 1, 
       lwd = c(1, 1))
```

Visual inspection of the series reveals a sustained upward trend over the 1990-2026 period, suggesting the presence of a stochastic trend. Furthermore, the fluctuations in the detrended series (shown on the right) exhibit time-varying volatility, indicating that the variance is not constant over time. These observations strongly suggest that the process is non-stationary. To rigorously confirm this assessment, we perform a comprehensive suite of statistical tests, including the Augmented Dickey-Fuller (ADF) and Phillips-Perron (PP) unit root tests, as well as the KPSS stationarity test.

#### ADF test

To rigorously assess the stationarity of the $ln(X_t)$ series, we follow a sequential testing procedure. We initially consider the most general specification, which includes both a linear deterministic trend and an intercept (the "trend" model):

\$\$\\Delta y_t = \\alpha + \\beta t + \\gamma y\_{t-1} + \\sum\_{i=1}\^p \\delta_i \\Delta y\_{t-i} + \\varepsilon_t\$\$

```{r, echo = FALSE}
#install.packages("tseries")
library(urca)
library(tseries)

#--- 1. TEST AUGMENTED DICKEY-FULLER (ADF) ---
adf_trend <- ur.df(lxt, type = "trend", lags = 12, selectlags = "BIC")
summary(adf_trend)

```

Where $y_t = ln(X_t)$. The results for this specification show that the test statistic for the unit root ($\tau_3 = -2.4846$) is greater than the 5% critical value (-3.42), leading to the non-rejection of the null hypothesis $H_0: \gamma = 0$. Furthermore, the joint test for a unit root and the absence of a deterministic trend ($\phi_3 = 3.5602$) is below the critical threshold (6.30).

This suggests that the deterministic trend $\beta t$ is not statistically significant. Consequently, to increase the power of the unit root test, we move to a more parsimonious specification by removing the trend component, known as the "drift" model:

\$\$\\Delta y_t = \\alpha + \\gamma y\_{t-1} + \\sum\_{i=1}\^p \\delta_i \\Delta y\_{t-i} + \\varepsilon_t\$\$

```{r, echo=FALSE}
# Modèle avec constante uniquement (Type = "drift")
adf_drift <- ur.df(lxt, type = "drift", lags = 12, selectlags = "BIC")
summary(adf_drift)


```

The test statistic for the unit root is $\tau_2 = -2.1667$, which remains above the 5% critical value of -2.87. Moreover, the joint test for the unit root and the intercept ($\phi_1 = 3.0132$) is below its critical threshold (4.61), indicating that the intercept $\alpha$ is not statistically significant. This leads us to the final and most parsimonious specification, the 'none' model, to confirm the presence of a pure unit root : \$\$\\Delta y_t = \\gamma y\_{t-1} + \\sum\_{i=1}\^p \\delta_i \\Delta y\_{t-i} + \\varepsilon_t\$\$

```{r, echo=FALSE}
# Modèle sans constante ni tendance (Type = "none")
adf_none <- ur.df(lxt, type = "none", lags = 12, selectlags = "BIC")
summary(adf_none)

```

Across all three specifications (trend, drift, and none), the null hypothesis of a unit root could not be rejected at the 5% level. Furthermore, the joint tests indicated that deterministic components were not statistically significant under the unit root hypothesis. We therefore conclude that the French Dairy Production Index is a **non-stationary process of order 1**, or $I(1)$.

To satisfy the stationarity requirement for $ARMA$ modeling, we apply a first-order differentiation to the logarithmic series: \$\$W_t = \\nabla ln(X_t) = ln(X_t) - ln(X\_{t-1})\$\$

#### Phillips-Perron test

```{r echo=FALSE}
pp_test <- ur.pp(lxt, type = "Z-tau", model = "constant", lags = "short") 
summary(pp_test)

```

#### KPSS test

```{r, echo=FALSE}
kpss_test <- ur.kpss(lxt, type="mu", lags="short", use.lag = 12)
summary(kpss_test)

```

To rigorously justify the transformation of the series, we performed a comprehensive suite of unit root and stationarity tests. The **Augmented Dickey-Fuller (ADF)** and **Phillips-Perron (PP)** tests (constant specification) both fail to reject the null hypothesis of a unit root at the 5% level (with statistics of $-2.16$ and $-2.70$ respectively). Furthermore, the **KPSS test** strongly rejects the null hypothesis of stationarity, with a test statistic of $2.6916$ far exceeding the 5% critical value of $0.463$. The convergence of these results definitively characterizes the $ln(X_t)$ series as non-stationary, specifically an $I(1)$ process. Consequently, we apply a first-order differentiation to stabilize the mean and achieve stationarity.

##### Table 1: Unit Root and Stationarity Tests Results for $ln(X_t)$

|  |  |  |  |  |  |
|----|----|----|----|----|----|
| **Test** | **Specification** | **Null Hypothesis (H0​)** | **Statistic** | **Critical Value (5%)** | **Result** |
| **ADF** | Trend, Intercept | Unit Root | $\tau_3 = -2.48$ | $-3.42$ | Fail to reject $H_0$ |
|  |  | Joint Test ($\gamma=0, \beta=0$) | $\phi_3 = 3.56$ | $6.30$ | Fail to reject $H_0$ |
| **ADF** | Intercept | Unit Root | $\tau_2 = -2.16$ | $-2.87$ | Fail to reject $H_0$ |
|  |  | Joint Test ($\gamma=0, \alpha=0$) | $\phi_1 = 3.01$ | $4.61$ | Fail to reject $H_0$ |
| **ADF** | None | Unit Root | $\tau_1 = 1.12$ | $-1.95$ | Fail to reject $H_0$ |
| **PP** | Intercept | Unit Root | $Z\text{-}\tau = -2.71$ | $-2.87$ | Fail to reject $H_0$ |
| **KPSS** | Trend | Stationarity | $LM = 2.6916$ | $0.463$ | **Reject** $H_0$ |

### c) Transform and visualize the new serie

```{r, echo=FALSE}
# Création de la série différenciée 
dlxt <- diff(lxt)

# Graphique comparatif (Question 3)
par(mfrow = c(1, 2)) 
plot(lxt, main = "Before", col = "blue", ylab = "Log-Index")
plot(dlxt, main = "After: First Difference", col = "gray", ylab = "Growth Rate")
abline(h = 0, lty = 2) # Ajoute une ligne à zéro pour voir la stationnarité en moyenne

lines(ma(dlxt, order = 12, centre = TRUE), col = "firebrick", lwd = 1) 

legend("bottomright", 
       legend = c("Log(Xt)", "Tendance"),
       col = c("gray", "firebrick"), 
       lty = 1, 
       lwd = c(1, 1))

```

The visual inspection of the first-differenced series $W_t = \nabla \ln(X_t)$ confirms the effectiveness of the transformation. The series now exhibits a clear mean-reverting behavior, oscillating stably around a zero mean with no discernible trend. Furthermore, the variance appears to be relatively constant over time, despite minor volatility clusters. These observations provide strong visual evidence that the differentiated series is stationary, fulfilling the necessary condition for $ARMA$ identification.

## Part II : ARMA models

### a) Choix du modèle ARMA et estimation des paramètres

```{r, echo=FALSE}
# Visualisation des corrélogrammes
library(ggplot2)
library(ggcorrplot)
par(mfrow = c(1, 2))
ggAcf(dlxt, lag.max = 80, plot = T)
ggPacf(dlxt, lag.max = 80, plot = T)
```

The identification of the optimal orders $p$ and $q$ is derived from the autocorrelation functions of the first-differenced series $\nabla \ln(X_t)$. The ACF plot reveals a single significant negative spike at the first lag ($k=1$), followed by an immediate truncation where subsequent correlations fall within the confidence intervals. This pattern is a classic signature of a first-order moving average process, justifying a maximum order of $q_{max} = 1$.

Simultaneously, the **PACF plot** displays significant coefficients for the first three lags before dropping sharply at lag 4. This truncation suggests that the internal memory of the series is effectively characterized by an autoregressive structure of order 3, setting the search limit at $p_{max} = 3$.

As a result, the model selection via grid search is restricted to the $ARIMA(p,1,q)$ family with the following constraints:

-   **Autoregressive order (**$p$): $p \leq 3$

-   **Moving average order (**$q$): $q \leq 1$

We are now going to test all these specifications, and select the model that has the lower information criteria AIC/BIC.

#### Estimation

```{r, echo =FALSE}

# Fonction qui qppliaue le test de Ljung Box aux résidus
Qtests <- function(series, k, fitdf=0) {
  pvals <- apply(matrix(1:k), 1, FUN=function(l) {
    pval <- if (l<=fitdf) NA else Box.test(series, lag=l, type="Ljung-Box", fitdf=fitdf)$p.value
    return(c("lag"=l,"pval"=pval))
  })
  return(t(pvals))
}

# fonction de test des significations individuelles des coefficients
signif <- function(estim){
  coef <- estim$coef
  se <- sqrt(diag(estim$var.coef))
  t <- coef/se
  pval <- (1-pnorm(abs(t)))*2
  return(rbind(coef,se,pval))
}


```

```{r, echo = FALSE}
# Entrainement des modèles candidats
model1= arima(dlxt,c(0,0,1), method = "ML")
model2=arima(dlxt,c(1,0,0), method = "ML")
model3=arima(dlxt,c(1,0,1), method = "ML")
model4=arima(dlxt,c(2,0,0), method = "ML")
model5=arima(dlxt,c(2,0,1), method = "ML")
model6=arima(dlxt,c(3,0,0), method = "ML")
model7=arima(dlxt,c(3,0,1), method = "ML")
```

```{r, echo=FALSE}
signif(model1)
signif(model2)
signif(model3)
signif(model4)
signif(model5)
signif(model6)
signif(model7)
```

Among all candidates, only ARMA(2,1) and ARMA(3,1) have coefficients that are non significant !

```{r, echo=FALSE}
Qtests(model1$residuals,24,1)
Qtests(model2$residuals,24,1)
Qtests(model3$residuals,24,2)
Qtests(model4$residuals,24,2)
Qtests(model6$residuals,24,3)
```

Only ARMA(1,1) passed the Ljung Box Test.

```{r echo=FALSE}
aic <- AIC(model1, model2, model3, model4, model5, model6, model7)
bic <- BIC(model1, model2, model3, model4, model5, model6, model7)
which.min(bic$BIC)
which.min(aic$AIC)
```

Finally, the best model is ARIMA(1,0,1). The equation is: \$\$\\ln(X_t) = 1,2651 \\ln(X\_{t-1}) - 0,2651 \\ln(X\_{t-2}) + 0,0004 + \\epsilon_t - 0,7237 \\epsilon\_{t-1}\$\$
